Combining information from photometric and spectroscopic data:\\ propogating uncertainties in spectrophotometric calibration

### 2015-05-12 tex/ms.tex
remove much of the introductory text, plotting for fully burned-in mocks

In probabilistic inference, all information flows from the data to the
parameters of interest by means of a likelihood function, or an
expression for the probability of the data conditioned on the
parameters.
If there are $N$ independent sources of data, for each of which there
is a likelihood function, the full-data likelihood is the product of
the $N$ individual likelihoods (or the log-likelihood is a sum of the
$N$ individual log-likelihoods).
There is no bespoke ``weighting'' of the data that is justifiable in
this framework.
This creates some apparent paradoxes!
One is the following:

Imagine you have three photometric bands through which you have
observed some object (independently, say), and also a 2500-pixel
spectrum, where the pixels are defined such that they are precisely or
approximately independent measurements.
For these purposes, two spectral pixels are very close to independent
measurements if they are derived from disjoint sets of detector pixels
in the spectrograph detector plane.
Each photometric band and each spectroscopic pixel has a
log-likelihood function, and these 2503 log-likelihood functions must
be (by the Rules of Righteousness) coadded equally into one full-data
log-likelihood.
The log-likelihood is therefore fully dominated by the spectral
pixels!
This despite the fact that in almost all astronomical investigations,
the photometry is far more reliable, per measurement, than the
spectra.
That is, if the spectral pixels tend to prefer one model, or one set
of parameters, and the photometry another, the spectra will almost
always ``win'' just by their overwhelming numbers.

What is a Righteous Probabilistic Investigator (RPI) to do?
One option---fully unjustified---is to ``down-weight'' the
log-likelihoods of the spectral pixels relative to the photometric
measurements.
This is equivalent to taking the likelihoods to some power, which has
no justification in inference (though it is sometimes used in
optimization and sampling as an ``annealing'' move).
This down-weighting is not justified, but there is even more
ad-hockery than this, because there is no principled way to choose the
unprincipled weights.

The Right Thing To Do is to express the \emph{reason} that the RPI
believes the spectroscopic data to be less reliable \emph{within} the
likelihood function.
That is, the RPI must add nuisance parameters to parameterize the
unreliability of the spectroscopy, and infer those along with the
parameters of interest.
If this parameterization is good, the information in the spectroscopy
data will not flow entirely to the parameters of interest; instead,
some will flow to the nuisance parameters.
This will reduce the importance of the spectroscopy relative to the
photometry and the RPI will be well pleased with the dominance of the
photometric data in the final inferences of the parameters of
interest.
This amounts to propogating the uncertainties in the
spectrophotometric calibration (or alternatively in the continuum
normalization).
In what follows, we provide an example of this kind of
parameterization and inference.

Our example involves instantiating a ``calibration adjustment'' model,
which itself has both a polynomial component and a non-parametric
component.
The parameters of the calibration adjustment are nuisances, from a
scientific standpoint.
However, because the calibration adjustment is inferred along with the
parameters of interest, the probabilistic inference that provides
parameter estimates \emph{also} provides a spectrophotometric
calibration of the spectral data.
That is, the method we present has the potential to \emph{obviate
spectrophotometric calibration}.

We find that this potential---obviation of calibration---is realized.
That is, if the procedures demonstrated below become widely adopted,
some or perhaps most astronomical spectroscopic projects could
potentially drop or dramatically reduce their spectrophotometric
calibration programs.
This would reduce telescope overheads and data-analysis time and
complexity, delivering economic and sanity benefits to the whole
community.

@@

Nebular emission lines (which may arise from HII regions around the
clusters or from diffuse ionized gas) and interstellar absorption
lines are modeled as a sum of $\delta$-functions of unknown amplitude
$a_\ell$, and with known central wavelengths $\lambda_\ell$, broadened
by the gas velocity profile, $\sigma_{gas}$.  For emission lines the
amplitudes are constrained to be positive, while for interstellar
absorption (e.g. Na D and Ca K) the amplitudes are constrained to be
negative. This framework can be easily extended to include more
complex models for the nebular emission, such as allowing line
dependent widths and velocities, and using Voight profiles for the
line shapes.

@@

In Figure \ref{fig:speconly_uncalibrated} we show the results of
inference on an \emph{uncalibrated} mock spectrum. {\color{red} This
  is not converged}

>>>>

Imagine you have three photometric bands through which you have
observed some object (independently, say), and also a 2500-pixel
spectrum, where the pixels are defined such that they are precisely or
approximately independent measurements.
For these purposes, two spectral pixels are very close to independent
measurements if they are derived from disjoint sets of detector pixels
in the spectrograph detector plane.
Each photometric band and each spectroscopic pixel has a
log-likelihood function, and these 2503 log-likelihood functions must
be coadded equally into one full-data
log-likelihood.
The log-likelihood is therefore fully dominated by the spectral
pixels!
This despite the fact that in almost all astronomical investigations,
the photometry is far more reliable, per measurement, than the
spectra.
That is, if the spectral pixels tend to prefer one model, or one set
of parameters, and the photometry another, the spectra will almost
always ``win'' just by their overwhelming numbers.

The Right Thing To Do is to express the \emph{reason} that the
spectroscopic data are less reliable \emph{within} the
likelihood function.
That is, the one must add nuisance parameters to parameterize the
unreliability of the spectroscopy, and infer those along with the
parameters of interest.
If this parameterization is good, the information in the spectroscopy
data will not flow entirely to the parameters of interest; instead,
some will flow to the nuisance parameters.
This will reduce the importance of the spectroscopy relative to the
photometry.
This amounts to propogating the uncertainties in the
spectrophotometric calibration (or alternatively in the continuum
normalization).
In what follows, we provide an example of this kind of
parameterization and inference.

Our example involves instantiating a ``calibration adjustment'' model,
which itself has both a polynomial component and a non-parametric
component.
The parameters of the calibration adjustment are nuisances, from a
scientific standpoint.
However, because the calibration adjustment is inferred along with the
parameters of interest, the probabilistic inference that provides
parameter estimates \emph{also} provides a spectrophotometric
calibration of the spectral data.

@@

Nebular emission lines (which may arise from HII regions around the
clusters or from diffuse ionized gas) and interstellar absorption
lines are modeled as a sum of $\delta$-functions of unknown amplitude
$a_\ell$, and with known central wavelengths $\lambda_\ell$, broadened
by the gas velocity profile, $\sigma_{gas}$.  For emission lines the
amplitudes are constrained to be positive, while for interstellar
absorption (e.g. NaD and CaK) the amplitudes are constrained to be
negative. This framework can be easily extended to include more
complex models for the nebular emission, such as allowing line
dependent widths and velocities, and using Voight profiles for the
line shapes.

@@

In Figure \ref{fig:speconly_uncalibrated} we show the results of
inference on an \emph{uncalibrated} mock spectrum.  The results are
similar to the case of the \emph{calibrated} spectrum.
